Fast Fourier transformation of electromagnetic data for computationally expensive kernels

### 2020-12-01 ManuscriptTime.tex
Clean diff

\title[\replaced{Transforming expensive EM kernels}{Time-domain CSEM}]{
  \replaced{Fast Fourier transformation of electromagnetic data for
  computationally expensive kernels}{Fast time-domain electromagnetic modelling
  in the frequency domain}}

@@

\begin{summary}
%
\deleted{Modelling three-dimensional time-domain controlled-source
electromagnetic data with frequen\-cy-domain solvers is an alternative to
time-domain solvers.} \added{Three-dimensional controlled-source
electromagnetic data is often computed directly in the domain of interest,
either in the frequency domain or in the time domain. Computing it in one
domain and transforming it via a Fourier transform to the other domain is a
viable alternative.} It requires the \replaced{evaluation}{computation} of many
\replaced{responses in the computational domain}{frequencies} if standard
Fourier transforms are used\replaced{. This}{, which} can make it prohibitively
expensive \replaced{if the kernel is time-consuming as is the case in
three-dimensional electromagnetic modelling}{in comparison with a time-domain
computation}. The speed of \replaced{such}{time-domain} modelling
\replaced{that is obtained through a transform}{with frequency-domain
computation} is defined by three key points: solver, method and implementation
of the Fourier transformation, and gridding.
%
The faster the \deleted{frequency-domain} solver, the faster
\deleted{time-domain} modelling will be. It is important that the solver is
robust over a wide range of \added{values (}frequencies\added{ or times)}.
\deleted{A three-dimensional, iterative, frequency-domain multigrid solver
satisfies this re\-quirement, but any other robust solver will do.}
%
The method should require as few \replaced{kernel evaluations}{frequencies} as
possible while remaining robust. As the frequency \added{and time}
range\added{s} span\deleted{s} many orders of magnitude, the required
\replaced{values}{frequencies} are ideally equally spaced on a logarithmic
scale. The \added{proposed} fast \deleted{frequency-to-time domain} method uses
either the digital linear filter method or the logarithmic fast Fourier
transform together with a careful selection of \replaced{evaluation
points}{frequencies} and interpolation. \replaced{In frequency-to-time domain
tests t}{T}his methodology requires typically 15 to 20 frequencies to cover a
wide range of offsets\deleted{in one-dimensional test models}.
%
The gridding should be frequency-\added{ or time-}dependent, which is
accomplished by making it a function of skin depth. Optimising for the least
number of required cells \deleted{for a given frequency} should be combined
with optimising for computational speed.
%
Looking carefully at these \deleted{three} points results in much smaller
computation times with speedup factors of ten or more over previous results. A
\deleted{frequency-domain} computation \replaced{followed by transformation}{of
time-domain data} can therefore be competitive to \replaced{computation in the
required }{time-}domain \deleted{solvers} if the required \replaced{evaluation
points}{frequencies} and the corresponding grids are carefully chosen.
%
\end{summary}

@@

The controlled-source electromagnetic (CSEM) method is one of the common
non-seismic tools in exploration geophysics, not only in hydrocarbon
exploration \citep{GEO.10.Constable}, but also in the search for sulfides
\citep{GRL.19.Gehrmann}, water \citep{GEO.05.Pedersen}, geothermal sources
\citep{WGC.15.Girard}, or geological purposes \citep{NAT.19.Johanson}. While
current sources with a few frequencies are used in the deep marine environment,
transient measurements are more common in the shallow marine environment and on
land \citep[e.g., ][]{GEO.07.Ziolkowski, SEG.07.Andreis, SEG.07.Avdeeva}. One
of the main reasons is the dominance of the airwave in shallow marine and
terrestrial measurements, which can be better separated in the time domain.
CSEM is usually divided into frequency- and time-domain methods, depending
whether the source signal is a continuous waveform, such as a sine, or a finite
waveform, such as a pseudo-random binary sequence (PRBS). A numerical
comparison of the two methods is given by \cite{GP.13.Conell}. Acquired CSEM
data are subsequently often analysed (modelled and inverted for) in their
respective domain, either the frequency or time domain. Modellers of layered
media usually exploit the horizontal shift-invariance by computing the
responses in the wavenumber-frequency domain followed by a 2D inverse spatial
Fourier transform, also called Hankel transform, to the space-frequency domain,
and a regular inverse Fourier transform if time-domain data are required
\citep[e.g., ][]{GEO.15.Hunziker}.

CSEM codes for 2D and 3D computations, on the other hand, mostly compute their
responses directly in the required domain, either frequency or time. There is a
wealth of 3D electromagnetic codes, with a first boom in the 1990's, with a
good state of the art provided by \cite{B.SEG.99.Oristaglio}. A second wave
occurred during the CSEM boom in the hydrocarbon industry in the early 21th
century, see, e.g., \cite{SG.05.Avdeev} and \cite{SG.10.Borner} for
comprehensive overviews. \replaced{Focusing on r}{R}ecent publications
\added{after these review articles} span\added{s} the width of numerical
methods from integral equation codes \citep{MGS.17.Kruglyakov} and differential
equation methods such as finite differences
\replaced{\mbox{\citep{GEO.10.Mittet, GEO.11.Zaslavsky, CAG.13.Sommer}}}
{\mbox{\citep{CAG.13.Sommer}}}, finite elements \citep{GJI.13.Grayver}, and
finite volumes \citep{GEO.14.Jahandari}. Besides the different methods there
are also many different types of discretisation. Commonly used, besides
rectilinear Yee grids \citep{IEEE.66.Yee} as used by our code, are Lebedev
grids \citep{CMMP.64.Lebedev} but also unstructured grids using, for instance,
tetrahedral \citep{CAG.17.Cai}. The computational requirement to model CSEM
data also depends heavily on the technique which is used to solve the system,
where there are direct solvers \citep{GEO.15.Grayver}, indirect solvers
\citep{GJI.15.Jaysaval}, and more recently also hybrid solvers
\citep{GEO.18.Liu}.

Under certain conditions it can be \replaced{a viable alternative}{competitive}
to model time-domain data with a frequency-domain code, as shown by
\cite{GEO.08.Mulder}. \added{They compared the computational complexity of
explicit time-stepping schemes (Du Fort-Frankel method as used in, e.g.,
\mbox{\cite{GEO.04.Commer, GEO.07.Maao}}); implicit schemes \mbox{\citep[e.g.,
][]{GEO.04.Haber}}, and matrix exponentials and Lanczos reduction schemes
\mbox{\cite[e.g., ][]{RS.94.Druskin}} with the Fourier transform of
frequency-domain solutions. Their conclusion was that the Fourier transform
method can be favourable if the number of required frequencies is small
relative to the number of grid points in each coordinate.} The conditions
\added{for fast computation of time-domain data with a frequency-domain code}
are: a sufficiently powerful solver, appropriate frequency selection and
interpolation, and an automated gridding, for which they used the
multi-frequency CSEM approach presented by \cite{GEO.07.Plessix}. We build upon
these results but improve the run time from hours to minutes. The main reasons
for this significant speed-up are an adaptive, frequency-dependent gridding
scheme that minimizes the required cells in each dimension, and a logarithmic
Fourier transform such as digital linear filters \citep[DLF, ][]{GP.71.Ghosh}
or the logarithmic fast Fourier transform \citep[FFTLog, ][]{RAS.00.Hamilton}
to go from the frequency to the time domain. The latter makes it also possible
to only use the imaginary part of the frequency-domain response, which has
advantages when it comes to interpolation.

\added{While we use the finite-integration technique
\mbox{\citep{AEU.77.Weiland}} in the frequency domain to show our method it is
important to note that our ideas can be used with any solver, it is not bound
to one or another frequency- or time-domain solver. The ideas for the frequency
selection can be applied to time selection, and time-dependent, adaptive
gridding exists as well \mbox{\citep[e.g., ][]{RS.06.Commer}}.} In the next
section, we briefly review the methodology as introduced by
\cite{GEO.07.Plessix} and \cite{GEO.08.Mulder} and highlight their advantages
and shortcomings. This is followed by an outline of our changes to the Fourier
transform and the adaptive gridding. Finally, we demonstrate the efficiency of
the approach with some numerical results.

@@

Being able to model CSEM data both in the frequency-domain and in the
time-domain can be desirable, as both domains have advantages and
disadvantages. One way to achieve this is to implement Maxwell's equation in
both domains\deleted{, as it is done, for instance, in \simpeg (Cockett et al.,
2015)}. Another approach is to have Maxwell's equation only implemented in one
domain, and use Fourier transforms to go to the other. However, this approach
can be costly, as many frequencies over a wide range are required to go from
the time domain to the frequency domain, or many times over a wide range for
the opposite direction. We present a methodology which significantly reduces
the amount and range of the required frequencies, translating into a
significant reduction in computation time. The required computation grids for
low frequencies (in our case in the order of 0.001\,Hz) and high frequencies
(around 100\,Hz) are hugely different. Low frequencies can be computed on a
coarser grid, but they require a much larger domain in order to avoid boundary
effects. High frequencies, on the other hand, require denser gridding, but they
are much more limited in reach. An adaptive gridding scheme is therefore
required. We build our approach upon \cite{GEO.07.Plessix}, who presented such
an adaptive gridding for multi-frequency (and multi-source) CSEM modelling.

An adaptive grid in the frequency domain is naturally based on the skin depth,
which is the distance after which the amplitude of the electromagnetic field is
decayed by $1/e\approx 37\,\%$. The skin depth $\delta$ is a function of
conductivity and frequency, and for the diffusive approximation of Maxwell's
equation in an isotropic, homogeneous medium is given by \citep[e.g.,][
equation~1.53]{B.SEG.88.Ward}
%
\begin{equation}
  \delta = \sqrt{\frac{2}{\omega\mu\sigma}}
  \ \approx \
  % \frac{503.3}{\sqrt{f\sigma}} \, ,
  % 503.3(f\sigma)^{-1/2} \, ,
  503.3/\sqrt{f\sigma} \, ,
  \label{eq:skindepth}
\end{equation}
%
where $\sigma$ is conductivity (S/m), $\omega=2\pi f$ is angular frequency of
frequency $f$ (Hz), and $\mu$ is magnetic permeability (H/m). The approximation
is obtained by using the free-space value of magnetic permeability,
$\mu_0=4\pi\times10^{-7}\,$H/m.

\cite{GEO.07.Plessix} define the minimum cell width $\Delta_\mathrm{min}$ as a
fraction of the minimum skin depth $\delta_\mathrm{min}$, where
$\delta_\mathrm{min}$ should be 2--3 times bigger than $\Delta_\mathrm{min}$.
The cells have to be smallest around the source; in the marine case, the
minimum skin depth is therefore defined by the conductivity of seawater.
However, this yields quite large cells for low frequencies, so special care has
to be taken around the source by defining a maximum allowed
$\Delta_\mathrm{min}$. The grid dimension, on the other hand, is defined as a
function of the skin depth for the average conductivity of the background,
$\delta_\mathrm{ave}$. They use four times $\delta_\mathrm{ave}$ for the $x$-,
$y$-, and downward $z$-directions, and a fixed 50\,km for the upward
$z$-direction to account for the airwave. To reduce the number of cells, it is
desirable to introduce stretching, at least in the boundary zone outside of the
area where source and receivers are located.

\cite{GEO.08.Mulder} provide a computational complexity analysis of various
methods to model transient electromagnetic responses directly in the
time-domain, and compare it to the computation of transient EM responses in the
frequency-domain with a subsequent Fourier transform. They conclude the review
by stating \emph{\guillemotleft Although it remains to be seen which of the
four methods requires the least computer time for a given accuracy, the
frequency-domain approach appears to be attractive.\guillemotright} Their
approach is to minimize the computation time by having, in addition to the just
introduced adaptive gridding, an adaptive frequency selection scheme. This
scheme starts with computing the responses for a set of just five frequencies,
regularly sampled on a log-scale, from minimum to maximum required frequency.
All the other frequencies are interpolated with a shape-preserving
piecewise-cubic Hermite interpolation \citep[PCHIP, ][]{SIAM.80.Fritsch}.
Testing the stability of the obtained response by removing a single
frequency-value at a time their scheme decides if more frequencies in-between
the already computed ones are required. In this way frequencies are only added
if required, hence if certain criteria of response stability are not met. While
this method is good and effective for a single offset, it loses all its
advantages if one tries to compute different offsets within one computation, as
each offset requires a different set of adaptive frequencies. Additionally, it
hampers the parallelization over frequencies.

We present improvements to both the adaptive gridding and the transform from
frequency domain to time domain, which makes time-domain modelling with a
frequency-domain code even more competitive in comparison with time-domain
codes. It is important to note that while we show the transform from frequency
to time domain, the conclusions regarding the Fourier transform can equally
well be applied to the transform from time to frequency domain data for a code
that solves Maxwell's equations in the time domain. A variation of our method
was implemented, for instance, by \custem \citep{GEO.19.Rochlitz} to obtain
frequency-domain data from time-domain computations (Rochlitz, private
communication, 2020).

@@

For the numerical computations we use the open-source (Apache License 2.0) code
\emg3d \citep{JOSS.19.Werthmuller}, a multigrid solver based on
\cite{GP.06.Mulder}, which can be used as a preconditioner for Krylov subspace
solvers or as a solver on its own. The multigrid approach works fine for the
diffusive approximation of Maxwell's equation, which assumes that
$\omega\varepsilon \ll \sigma$, where $\varepsilon$ is electric permittivity
(F/m). The remaining system to solve in the frequency domain is then given by
the second-order differential equation of the electric field,
%
\begin{equation}
    \mathrm{i}\omega\sigma \mathbf{E} +
    \nabla \times \mu^{-1} \nabla \times \mathbf{E}
    = -\mathrm{i}\omega\mathbf{J}_\mathrm{s} \, ,
  \label{eq:maxwell}
\end{equation}
%
where $\mathbf{E}$ is the electric field (V/m) and $\mathbf{J}_\mathrm{s}$ the
current source (A/m$^2$); time dependence is $\exp(\mathrm{i}\omega t)$. The
standard multigrid approach fails for severe stretching or strong anisotropy,
for which \deleted{\emg3d has implemented }known improvements such as
line-relaxation and semicoarsening \citep{ECCFD.06.Jonsthovel} \added{are
implemented}, with a non-standard Cholesky decomposition to speed up the
computation \citep{GEO.08.Mulder}. One of the big advantages of the multigrid
method is that it scales linearly (optimal) with the grid size in both CPU and
RAM usage \citep{B.Springer.20.Mulder}. This makes it feasible to run even big
models on standard computers, without the need for big clusters. All the
examples in this article are run on a laptop with an i7-6600U\,CPU\,@\,2.6\,GHz
(x4) and 16\,GB of memory, using Ubuntu \replaced{20.04}{18.04} and Python
\replaced{3.8}{3.7}. \added{All examples were run using a single thread, but
parallelization over frequencies is straight forward, as the computations are
independent.} For the semi-analytical computations of layered models we use the
open-source (Apache License 2.0) code \empymod \citep{GEO.17.Werthmuller}.

\section{Frequency selection} % % % % % % % % % % % % % % % % % % % % % % % % %

An important factor in terms of speed and accuracy for time-domain responses
obtained from frequency-domain computations is the selection of the required
frequencies. The fewer frequencies required, the quicker we obtain the
time-domain result. \added{The same applies for the reverse operation, the
fewer times required, the quicker we obtain frequency-domain results. Even
though we limit our analysis here to frequency-to-time transformation, most of
the arguments are reversible and applicable to time-to-frequency
transformations.} We decided to use a regular spacing of frequencies on a
log-scale, rather than an adaptive scheme. This approach is favourable if a
wide range of offsets is needed, as the required frequencies change with offset
and an adaptive frequency selection is therefore often tailored to a single
offset. Also, it allows for straightforward parallelization over frequencies,
which is not completely possible with an adaptive scheme.

For the actual transform we use either the digital linear filter (DLF) method
or the logarithmic fast Fourier transform (FFTLog). The DLF method was
introduced to geophysics by \cite{GP.71.Ghosh}, and is arguably the most common
method in EM geophysics for both its simplicity and its speed. It is
implemented for the Hankel and Fourier transforms in most EM modelling codes,
e.g., \cite{GEO.09.Key}. A simple tool to design digital linear filters was
recently presented by \cite{GEO.19.Werthmuller}, together with a comprehensive
overview of the history and development of DLF in geophysics. FFTLog,
introduced by \cite{RAS.00.Hamilton}, is another transform algorithm which
proved to be powerful for the frequency-to-time domain transformation of EM
responses, e.g., \cite{INT.14.Werthmuller}. In our tests they are both about
equal in speed and accuracy. DLF requires a wider range and many more
frequencies than the FFTLog. Both methods share some important characteristics
in comparison with the standard FFT: the required input frequencies are equally
spaced on a logarithmic scale\added{ (natural logarithm in the case of the DLF
and decimal logarithm in the case of the FFTLog)}, and they only require either
the real or the imaginary part of the frequency-domain response. We can take
advantage of that by using only the imaginary part of the frequency-domain
response. The imaginary part goes to zero when the frequency goes
\added{either} to zero or to infinity, with the advantage that knowing the
endpoints makes it possible to convert the extrapolation of missing frequencies
into \added{an }interpolation.

@@

The \added{shown} example model is a marine scenario with 1\,km water depth of
resistivity $\rho=0.3\,\ohmm$ ($\rho = \sigma^{-1}$) and a 100\,m thick target
of 100\,\ohmm at 1\,km below the seafloor in a background of 1\,\ohmm. The
source is 50\,m above the seafloor, the receiver is on the seafloor, and the
response is the inline $x$-directed $E$-field. The left subplot shows the
imaginary part in the frequency domain and the right plot the corresponding
impulse response in the time domain. The red lines are the
\replaced{semi-analytical responses}{precise results obtained from \empymod}.
The blue circles indicate the actually computed responses and the black dots
the frequencies which are interpolated or set to zero. The resulting
time-domain response has a relative error of less than 1\,\% everywhere except
for very early times.

>>>>

\title[Transforming expensive EM kernels]{Fast Fourier transformation of
electromagnetic data for computationally expensive kernels}

@@

\begin{summary}
%
Three-dimensional controlled-source electromagnetic data is often computed
directly in the domain of interest, either in the frequency domain or in the
time domain. Computing it in one domain and transforming it via a Fourier
transform to the other domain is a viable alternative. It requires the
evaluation of many responses in the computational domain if standard Fourier
transforms are used. This can make it prohibitively expensive if the kernel is
time-consuming as is the case in three-dimensional electromagnetic modelling.
The speed of such modelling that is obtained through a transform is defined by
three key points: solver, method and implementation of the Fourier
transformation, and gridding.
%
The faster the solver, the faster modelling will be. It is important that the
solver is robust over a wide range of values (frequencies or times). The method
should require as few kernel evaluations as possible while remaining robust. As
the frequency and time ranges span many orders of magnitude, the required
values are ideally equally spaced on a logarithmic scale. The proposed fast
method uses either the digital linear filter method or the logarithmic fast
Fourier transform together with a careful selection of evaluation points and
interpolation. In frequency-to-time domain tests this methodology requires
typically 15 to 20 frequencies to cover a wide range of offsets.
%
The gridding should be frequency- or time-dependent, which is accomplished by
making it a function of skin depth. Optimising for the least number of required
cells should be combined with optimising for computational speed. Looking
carefully at these points results in much smaller computation times with
speedup factors of ten or more over previous methods. A computation in one
domain followed by transformation can therefore be an alternative to
computation in the other domain domain if the required evaluation points and
the corresponding grids are carefully chosen.
%
\end{summary}

@@

The controlled-source electromagnetic (CSEM) method is one of the common
non-seismic tools in exploration geophysics, not only in hydrocarbon
exploration \citep{GEO.10.Constable}, but also in the search for sulfides
\citep{GRL.19.Gehrmann}, water \citep{GEO.05.Pedersen}, geothermal sources
\citep{WGC.15.Girard}, or for geological purposes \citep{NAT.19.Johanson}.
While current sources with a few frequencies are used in the deep marine
environment, transient measurements are more common in the shallow marine
environment and on land \citep[e.g., ][]{GEO.07.Ziolkowski, SEG.07.Andreis,
SEG.07.Avdeeva}. One of the main reasons is the dominance of the airwave in
shallow marine and terrestrial measurements, which can be better separated in
the time domain. CSEM is usually divided into frequency- and time-domain
methods, depending on whether the source signal is a continuous waveform, such
as a sine, or a finite waveform, such as a pseudo-random binary sequence
(PRBS). A numerical comparison of the two methods is given by
\cite{GP.13.Conell}. Acquired CSEM data are subsequently often analysed
(modelled and inverted for) in their respective domain, either the frequency or
time domain. Modellers of layered media usually exploit the horizontal
shift-invariance by computing the responses in the wavenumber-frequency domain
followed by a 2D inverse spatial Fourier transform, also called Hankel
transform, to the space-frequency domain, and a regular inverse Fourier
transform if time-domain data are required \citep[e.g., ][]{GEO.15.Hunziker}.

CSEM codes for 2D and 3D computations, on the other hand, mostly compute their
responses directly in the required domain, either frequency or time. There is a
wealth of 3D electromagnetic codes, with a first boom in the 1990's;
\cite{B.SEG.99.Oristaglio} provide a good review of the state of art at that
time. A second wave occurred during the CSEM boom in the hydrocarbon industry
in the early 21th century, see, e.g., \cite{SG.05.Avdeev} and
\cite{SG.10.Borner} for comprehensive overviews. Recent publications after
these review articles span the width of numerical methods from integral
equation codes \citep{MGS.17.Kruglyakov} and differential equation methods such
as finite differences \citep{GEO.10.Mittet, GEO.11.Zaslavsky, CAG.13.Sommer},
finite elements \citep{GJI.13.Grayver}, and finite volumes
\citep{GEO.14.Jahandari}. Besides the different methods there are also many
different types of discretisation. Commonly used, besides rectilinear Yee grids
\citep{IEEE.66.Yee} as used by our code, are Lebedev grids
\citep{CMMP.64.Lebedev} but also unstructured grids using, for instance,
tetrahedra \citep{CAG.17.Cai}. The computational requirement to model CSEM data
also depends heavily on the technique used to solve the system. There are
direct solvers \citep{GEO.15.Grayver}, indirect solvers
\citep{GJI.15.Jaysaval}, and more recently also hybrid solvers
\citep{GEO.18.Liu}.

Under certain conditions it can be a viable alternative to model time-domain
data with a frequency-domain code, as shown by \cite{GEO.08.Mulder}. They
compared the computational complexity of explicit time-stepping schemes (Du
Fort-Frankel method as used in, e.g., \cite{GEO.04.Commer, GEO.07.Maao});
implicit schemes \citep[e.g., ][]{GEO.04.Haber}, and matrix exponentials and
Lanczos reduction schemes \cite[e.g., ][]{RS.94.Druskin} with the Fourier
transform of frequency-domain solutions. Their conclusion was that the Fourier
transform method can be favourable if the number of required frequencies is
small relative to the number of grid points in each coordinate. The conditions
for fast computation of time-domain data with a frequency-domain code are: a
sufficiently powerful solver, appropriate frequency selection and
interpolation, and an automated gridding, for which they used the
multi-frequency CSEM approach presented by \cite{GEO.07.Plessix}. We build upon
these results but improve the run time from hours to minutes. The main reasons
for this significant speed-up are an adaptive, frequency-dependent gridding
scheme that minimizes the required cells in each dimension, and a logarithmic
Fourier transform such as digital linear filters \citep[DLF, ][]{GP.71.Ghosh}
or the logarithmic fast Fourier transform \citep[FFTLog, ][]{RAS.00.Hamilton}
to go from the frequency to the time domain. The latter makes it also possible
to only use the imaginary part of the frequency-domain response, which has
advantages when it comes to interpolation.

While we use the finite-integration technique \citep{AEU.77.Weiland} in the
frequency domain to demonstrate our method, it is important to note that our
ideas can be used with any solver and is not bound to one or another frequency-
or time-domain solver. The ideas for the frequency selection can be applied to
time selection, and time-dependent, adaptive gridding exists as well
\citep[e.g., ][]{RS.06.Commer}. In the next section, we briefly review the
methodology as introduced by \cite{GEO.07.Plessix} and \cite{GEO.08.Mulder} and
highlight their advantages and shortcomings. This is followed by an outline of
our changes to the Fourier transform and the adaptive gridding. Finally, we
demonstrate the efficiency of the approach with some numerical results.

@@

Being able to model CSEM data both in the frequency domain and in the time
domain can be desirable, as both domains have advantages and disadvantages. One
way to achieve this is to implement Maxwell's equations in both domains.
Another approach is to have Maxwell's equations only implemented in one domain,
and use Fourier transforms to go to the other. However, this approach can be
costly, as many frequencies over a wide range are required to go from the
frequency domain to the time domain, or many times over a wide range for the
opposite direction. We present a methodology which significantly reduces the
amount and range of the required frequencies, translating into a significant
reduction in computation time. The required computation grids for low
frequencies (in our case in the order of 0.001\,Hz) and high frequencies
(around 100\,Hz) are hugely different. Low frequencies can be computed on a
coarser grid, but they require a much larger domain in order to avoid boundary
effects. High frequencies, on the other hand, require denser gridding, but they
are much more limited in reach. An adaptive gridding scheme is therefore
required. We build our approach upon \cite{GEO.07.Plessix}, who presented such
an adaptive gridding for multi-frequency (and multi-source) CSEM modelling.

An adaptive grid in the frequency domain is naturally based on the skin depth,
which is the distance after which the amplitude of the electromagnetic field is
decayed by $1/e\approx 37\,\%$. The skin depth $\delta$ is a function of
conductivity and frequency, and for the diffusive approximation of Maxwell's
equations in an isotropic, homogeneous medium is given by \citep[e.g.,][
equation~1.53]{B.SEG.88.Ward}
%
\begin{equation}
  \delta = \sqrt{\frac{2}{\omega\mu\sigma}}
  \ \approx \ 503.3/\sqrt{f\sigma} \, ,
  \label{eq:skindepth}
\end{equation}
%
where $\sigma$ is conductivity (S/m), $\omega=2\pi f$ is angular frequency of
frequency $f$ (Hz), and $\mu$ is magnetic permeability (H/m). The approximation
is obtained by using the free-space value of magnetic permeability,
$\mu_0=4\pi\times10^{-7}\,$H/m.

\cite{GEO.07.Plessix} define the minimum cell width $\Delta_\mathrm{min}$ as a
fraction of the minimum skin depth $\delta_\mathrm{min}$, where
$\delta_\mathrm{min}$ should be 2--3 times bigger than $\Delta_\mathrm{min}$.
The cells have to be smallest around the source; in the marine case, the
minimum skin depth is therefore defined by the conductivity of seawater.
However, this yields quite large cells for low frequencies, so special care has
to be taken around the source by defining a maximum allowed
$\Delta_\mathrm{min}$. The grid dimension, on the other hand, is defined as a
function of the skin depth for the average conductivity of the background,
$\delta_\mathrm{ave}$. They use four times $\delta_\mathrm{ave}$ for the $x$-,
$y$-, and downward $z$-directions, and a fixed 50\,km for the upward
$z$-direction to account for the airwave. To reduce the number of cells, it is
desirable to introduce stretching, at least in the buffer zone outside of the
area where source and receivers are located.

\cite{GEO.08.Mulder} provide a computational complexity analysis of various
methods to model transient electromagnetic responses directly in the time
domain, and compare it to the computation of transient EM responses in the
frequency domain with a subsequent Fourier transform. They conclude the review
by stating \emph{\guillemotleft Although it remains to be seen which of the
four methods requires the least computer time for a given accuracy, the
frequency-domain approach appears to be attractive.\guillemotright} Their
approach is to minimize the computation time by having, in addition to the just
introduced adaptive gridding, an adaptive frequency selection scheme. This
scheme starts with computing the responses for a set of just five frequencies,
regularly sampled on a log-scale, from minimum to maximum required frequency.
All the other frequencies are interpolated with a shape-preserving
piecewise-cubic Hermite interpolation \citep[PCHIP, ][]{SIAM.80.Fritsch}.
Testing the stability of the obtained response by removing a single
frequency-value at a time their scheme decides if more frequencies in-between
the already computed ones are required. In this way frequencies are only added
if required, hence if certain criteria of response stability are not met. While
this method is good and effective for a single offset, it loses all its
advantages if one tries to compute different offsets within one computation, as
each offset requires a different set of adaptive frequencies. Additionally, it
hampers the parallelization over frequencies.

We present improvements to both the adaptive gridding and the transform from
frequency domain to time domain, which makes time-domain modelling with a
frequency-domain code viable in comparison with time-domain codes. It is
important to note that while we show the transform from frequency to time
domain, the conclusions regarding the Fourier transform can equally well be
applied to the transform from time-domain to frequency-domain data for a code
that solves Maxwell's equations in the time domain. A variation of our method
was implemented, for instance, by \custem \citep{GEO.19.Rochlitz} to obtain
frequency-domain data from time-domain computations (Rochlitz, private
communication, 2020).

@@

For the numerical computations we use the open-source (Apache License 2.0) code
\emg3d \citep{JOSS.19.Werthmuller}, a multigrid solver based on
\cite{GP.06.Mulder}, which can be used as a preconditioner for Krylov subspace
solvers or as a solver on its own. The multigrid approach works fine for the
diffusive approximation of Maxwell's equations, which assumes that
$\omega\varepsilon \ll \sigma$, where $\varepsilon$ is electric permittivity
(F/m). The remaining system to solve in the frequency domain is then given by
the second-order differential equation of the electric field,
%
\begin{equation}
    \mathrm{i}\omega\sigma \mathbf{E} +
    \nabla \times \mu^{-1} \nabla \times \mathbf{E}
    = -\mathrm{i}\omega\mathbf{J}_\mathrm{s} \, ,
  \label{eq:maxwell}
\end{equation}
%
where $\mathbf{E}$ is the electric field (V/m) and $\mathbf{J}_\mathrm{s}$ the
current source (A/m$^2$); time dependence is $\exp(\mathrm{i}\omega t)$. The
standard multigrid approach fails for severe stretching or strong anisotropy,
for which known improvements such as line-relaxation and semicoarsening
\citep{ECCFD.06.Jonsthovel} are implemented, with a non-standard Cholesky
decomposition to speed up the computation \citep{GEO.08.Mulder}. One of the big
advantages of the multigrid method is that it scales linearly (optimal) with
the grid size in both CPU and RAM usage \citep{B.Springer.20.Mulder}. This
makes it feasible to run even big models on standard computers, without the
need for big clusters. All the examples in this article are run on a laptop
with an i7-6600U\,CPU\,@\,2.6\,GHz (x4) and 16\,GB of memory, using Ubuntu
20.04 and Python 3.8. All examples were run using a single thread, but
parallelization over frequencies is straight forward, as the computations are
independent. For the semi-analytical computations of layered models we use the
open-source (Apache License 2.0) code \empymod \citep{GEO.17.Werthmuller}.

\subsection{Frequency selection}  % % % % % % % % % % % % % % % % % % % % % % %

An important factor in terms of speed and accuracy for time-domain responses
obtained from frequency-domain computations is the selection of the required
frequencies. The fewer frequencies required, the quicker we obtain the
time-domain result. The same applies for the reverse operation, the fewer times
required, the quicker we obtain frequency-domain results. Even though we limit
our analysis here to frequency-to-time transformations, most of the arguments
are reversible and applicable to time-to-frequency transformations. We decided
to use a regular spacing of frequencies on a log-scale, rather than an adaptive
scheme. This approach is favourable if a wide range of offsets is needed, as
the required frequencies change with offset and an adaptive frequency selection
is therefore often tailored to a single offset. Also, it allows for
straightforward parallelization over frequencies, which is not completely
possible with an adaptive scheme.

For the actual transform we use either the digital linear filter (DLF) method
or the logarithmic fast Fourier transform (FFTLog). The DLF method was
introduced to geophysics by \cite{GP.71.Ghosh}, and is arguably the most common
method in EM geophysics for both its simplicity and its speed. It is
implemented for the Hankel and Fourier transforms in most EM modelling codes,
e.g., \cite{GEO.09.Key}. A simple tool to design digital linear filters was
recently presented by \cite{GEO.19.Werthmuller}, together with a comprehensive
overview of the history and development of DLF in geophysics. FFTLog,
introduced by \cite{RAS.00.Hamilton}, is another transform algorithm which
proved to be powerful for the frequency-to-time domain transformation of EM
responses, e.g., \cite{INT.14.Werthmuller}. In our tests they are both about
equal in speed and accuracy. DLF requires a wider range and many more
frequencies than the FFTLog. Both methods share some important characteristics
in comparison with the standard FFT: the required input frequencies are equally
spaced on a logarithmic scale (natural logarithm in the case of the DLF and
decimal logarithm in the case of the FFTLog), and they only require either the
real or the imaginary part of the frequency-domain response. We can take
advantage of that by using only the imaginary part of the frequency-domain
response. The imaginary part goes to zero when the frequency goes either to
zero or to infinity, with the advantage that knowing the endpoints makes it
possible to convert the extrapolation of missing frequencies into an
interpolation.

@@

The shown example model is a marine scenario with 1\,km water depth of
resistivity $\rho=0.3\,\ohmm$ ($\rho = \sigma^{-1}$) and a 100\,m thick target
of 100\,\ohmm at 1\,km below the seafloor in a background of 1\,\ohmm. The
source is 50\,m above the seafloor, the receiver is on the seafloor, and the
response is the inline $x$-directed $E$-field. The left subplot shows the
imaginary part in the frequency domain and the right plot the corresponding
impulse response in the time domain. The red lines are the semi-analytical
responses. The blue circles indicate the actually computed responses and the
black dots the frequencies which are interpolated or set to zero. The resulting
time-domain response has a relative error of less than 1\,\% everywhere except
for very early times.

Figure \ref{fig:GUI-FFTLog-log} shows exactly the same on a logarithmic scale,
without the interactive GUI-related elements. It can be seen that with the
chosen frequencies the time-domain response starts to divert above about 100\,s
and below 0.3\,s. It also shows the oscillating high-frequency part, which is
hard to interpolate and we therefore set it to zero. Figure
\ref{fig:GUI-DLF-lin} is the same as Figure \ref{fig:GUI-FFTLog-lin}, but
transformed with the DLF method applying the 81-point sine-cosine filter from
\cite{GEO.09.Key}. The same frequencies were computed as in the FFTLog case and
the missing ones interpolated. The error of the corresponding time-domain
response is comparable, so either FFTLog or DLF can be used.
%
\begin{figure*}
  \centering
  \subfloat[]{{\includegraphics[width=0.5\fwidth]{02a-GUI-FFTLog-log}}%
  \label{fig:GUI-FFTLog-log}}\hfill
  \subfloat[]{{\includegraphics[width=0.5\fwidth]{02b-GUI-DLF-lin}}%
  \label{fig:GUI-DLF-lin}}\\
  \caption{(a) Same as Figure \ref{fig:GUI-FFTLog-lin}, but on a logarithmic
  scale. (b) Using DLF instead of FFTLog, but computing the same frequency
  range as for the FFTLog. The resulting time-domain response has comparable
  accuracy (compare the error to the error in Figure
  \ref{fig:GUI-FFTLog-lin}).}
  \label{fig:GUI-log-lin}
\end{figure*}
%
In the above example we used 20 frequencies, but that many would not be
required for the shown offset of 5\,km, a few of the lower and higher
frequencies could be left out. With this frequency-selection we can model a
wide range of offsets. This is shown in Figure \ref{fig:multi-offset}, where
the same kernels were used to yield the responses at offsets $r=1.5,3,6$, and
12\,km. We only need to compute 50\,\% of the required frequencies in the case
of the FFTLog, and only 14\,\% in the case of the DLF.
%
\begin{figure*}
  \centering
  \includegraphics[width=0.75\fwidth]{03-multi-offset}
  \caption{Normalized (a) frequency- and (b) time-domain responses for offsets
    $r=1.5,3,6$, and 12\,km using the parameters defined in Figure
    \ref{fig:GUI-FFTLog-lin}. The coloured symbols are the actually computed
    responses, the black dots are the responses which are set to zero (high
    frequencies) or interpolated (low frequencies). The grey curves are the
    analytical responses.}
  \label{fig:multi-offset}
\end{figure*}
%
